Unsupervised training of maximum-entropy models for lexical selection in rule-based machine translation

### 2015-02-09 2015-eamt-melsa/2015-eamt-melsa.tex
fix up some formatting issues
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@58777 72bbbca6-d526-0410-a7d9-f06f51895060

\author{
 Francis M.\ Tyers \\
 HSL-fakultehta, \\
 UiT Norgga \'{a}rktala\v{s} universitehta, \\
 N-9018 Romsa \\
 \and
 Felipe S\'anchez-Mart\'inez \\
 Dept.\ Lleng.\ i Sist.\ Inform., \\
 Universitat d'Alacant,\\
 E-03071 Alacant\\
 \and
 Mikel L.\ Forcada \\
 Dept.\ Lleng.\ i Sist.\ Inform., \\
 Universitat d'Alacant,\\
 E-03071 Alacant\\
}

@@

\begin{table}
 \begin{center}
  \begin{tabular}{|c|l|r|}
    \hline
       & \textbf{Sentence} & $p(g_i|S)$ \\
    \hline % Los pescados grandes peque\~{n}o come
     \(S_1\) & \emph{Arrain handiak txikia jaten du}  &  \\
                        $\tau(g_{11},S_1)$ & El pez grande se come el peque\~{n}o &  0.830 \\
                        $\tau(g_{12},S_1)$ & El pescado grande se come el peque\~{n}o & 0.134 \\
                        $\tau(g_{13},S_1)$ & El pescado capaz se come el peque\~{n}o & 0.034\\
                        $\tau(g_{14},S_1)$ & El pez capaz se come el peque\~{n}o  & 0.001\\
    \hline % El padre el pescado nos ha preparado para cenar
     \(S_2\) & \emph{Aitak arraina prestatu digu afaltzeko} &  \\
                        $\tau(g_{21},S_2)$ & Padre nos ha preparado pescado para cenar & 0.846 \\
                        $\tau(g_{22},S_2)$ & Pap\'{a} nos ha preparado pescado para cenar & 0.134 \\
                        $\tau(g_{23},S_2)$ & Padre nos ha preparado pez para cenar & 0.015 \\
                        $\tau(g_{24},S_2)$ & Pap\'{a} nos ha preparado pez para cenar & 0.005 \\
    \hline % Aqu\'{\i} el pescado muy es dulce
    $S_3$  & \emph{Hemen arraina oso goxoa da}  & \\
                        $\tau(g_{31},S_3)$ & Aqu\'{\i} el pescado es muy rico & 0.912 \\
                        $\tau(g_{32},S_3)$ & Aqu\'{\i} el pescado es muy suave & 0.067 \\
                        $\tau(g_{33},S_3)$ & Aqu\'{\i} el pez es muy rico & 0.011 \\
                        $\tau(g_{34},S_3)$ & Aqu\'{\i} el pez es muy suave & 0.008 \\
                        $\tau(g_{35},S_3)$ & Aqu\'{\i} el pez es muy dulce & 0.001 \\
                        $\tau(g_{36},S_3)$ & Aqu\'{\i} el pescado es muy dulce & 0.001 \\
    \hline % Un pescado grande del mar es
    \(S_4\) & \emph{Itsasoko arrain handi bat da} & \\
                        $\tau(g_{41},S_4)$ & Es un pez grande del mar & 0.595 \\
                        $\tau(g_{42},S_4)$ & Es un pescado grande del mar & 0.403 \\
                        $\tau(g_{43},S_4)$ & Es un pez capaz del mar & 0.001 \\
                        $\tau(g_{44},S_4)$ & Es un pescado capaz del mar & 0.001 \\
    \hline % El pescado grande por esos serv\'{\i}an para alimentar
    $S_5$ &  \emph{Arrain handiak horiez baliatzen ziren elikatzeko} &  \\
                        $\tau(g_{51},S_5)$ & Los peces grandes se nutr\'{\i}an de esos & 0.862 \\
                        $\tau(g_{52},S_5)$ & Los pescados grandes se nutr\'{\i}an de esos & 0.121 \\
                        $\tau(g_{53},S_5)$ & Los peces grandes se alimentaban de esos & 0.012 \\
                        $\tau(g_{54},S_5)$ & Los peces capaces se alimentaban de esos & 0.001 \\
                        $\tau(g_{55},S_5)$ & Los pescados grandes se alimentaban de esos & 0.001 \\
                        $\tau(g_{56},S_5)$ & Los pescados capaces se alimentaban de esos & 0.001 \\
                        $\tau(g_{57},S_6)$ & Los peces capaces se nutr\'{\i}an de esos & 0.001 \\
                        $\tau(g_{58},S_7)$ & Los pescados capaces se nutr\'{\i}an de esos & 0.001 \\
    \hline

  \end{tabular}
  \caption{A sample of source-language sentences, the output of the
    possible lexical-selection paths after scoring on a
    target-language model. The scores are normalised as fractional
    counts, which are used as inputs to the maximum-entropy training
    process. In all sentences, the resultant translation with the highest share of the 
    probability mass is the most adequate.}
   \label{tb:workedexample}
 \end{center}
\end{table}

@@

\begin{table}
  \centering
  \begin{tabular}{|l|r|r|r|r|r|}
    \hline
    \multirow{2}{*}{\textbf{Pair}} & \textbf{Lexical} & \multirow{2}{*}{\textbf{Ambig.}} & \textbf{Na\"{i}ve}& \textbf{Total} & \textbf{Ambig. of}\\\
                          & \textbf{items} &                            & \textbf{coverage} & \textbf{ambig.} & \textbf{ambig. words}\\
    \hline
    \texttt{br-fr} & 27,690 & 235  &  94.47 \%  & 1.02 & 2.5 \\
    \texttt{mk-en} & 33,229 & 1,323 & 92.17 \%  & 1.37 & 2.8 \\
    \texttt{eu-es} & 19,323 & 455 & 91.70 \% & 1.08 & 2.4 \\
    \texttt{en-es} & 30,186 & 822 & 98.08 \% & 1.05 & 2.2 \\
    \hline
  \end{tabular}
  \caption{Statistics about development status of language pairs. The second
    column gives the number of entries in the bilingual dictionaries of the systems;
    the third column gives the total number of source language words with more 
    than one translation in the target language; the fourth column gives the coverage
    of the dictionaries on the test corpus (see section~\ref{sec:eval-corpus}); the fifth
    column gives the total mean ambiguity for open classes (nouns, verbs, adjectives);
    and the final column gives the mean ambiguity for ambiguous words in open classes.}
  \label{table:lang-pair-stats}
\end{table}

@@

\label{sec:eval-metrics}
\subsubsection{Lexical-selection performance}
This is an intrinsic module-based evaluation of the performance of the
lexical-selection module.  It measures how well the lexical-selection
module disambiguates the output of the lexical-transfer module as
compared to a gold-standard corpus. For this task, we define a metric,
the lexical-selection error rate ({\sc ler}),
% see section~\ref{sec:ler}. 
\begin{figure}
\begin{center}
\setlength\tabcolsep{5.2pt}%
 \begin{tabular}{|l|llllll|}
   \hline
   \multicolumn{7}{|c|}{L'estiu \'{e}s una estaci\'{o} llarga}\\
   \hline
   $S$        & \emph{el} & \emph{estiu} & \emph{ser} & \emph{un} & \emph{estaci\'{o}} & \emph{llarg} \\
   \hline
   $T_s(s_i)$ & \{the\} & \{summer\} & \{be\} & \{a\} & \{station, season\} &  \{long, lengthy\} \\
   \hline 
   $T_r(s_i)$ & \{the\} & \{summer\} & \{be\} & \{a\} &  \{season\} & \{long\} \\
   \hline
   $T_t(s_i)$ & \{the\} & \{summer\} & \{be\} & \{a\} &  \{station\} & \{long\} \\
   \hline
 \end{tabular}
 \caption{An example input sentence in Catalan and the three sets of
   English translations used for calculating the lexical-selection
   error rate. The source sentence $S = (s_1, s_2, \ldots, s_{|S|})$
   has two ambiguous words, \emph{estaci\'{o}} and \emph{llarg}. There is one difference
   between the reference set $T_r(s_i)$ and the test set $T_t(s_i)$ of
   translations; as a result, the error rate for this sentence is
   50\%.}
\label{fig:ler-example}
\end{center}
\end{figure}

@@

\begin{table}
\begin{center}
 \begin{tabular}{|l|r|r||r|r|r|r|}
    \hline  
    %Pair           & Lines & Words & Extracted & \texttt{train} & \texttt{test} & \texttt{dev} & No. ambig & Av. ambig \\ 
    {\bf Pair}      & {\bf Lines} & {\bf Extract.} & \texttt{train} & \texttt{other} & {\bf No. amb} & {\bf Av. amb} \\ 
    \hline  
     %\texttt{br-fr} & 57,305  & 702,328  & 4,668 & 2,668 & 1,000  & 1,000 & 603  & 1.07 \\
     \texttt{br-fr} & 57,305  & 4,668 & 2,668 & 2,000  & 603  & 3.06 \\
    \hline  
     %\texttt{mk-en} & 190,493  & 4,259,338  & 19,747  & 17,747  & 1,000 & 1,000 & 13,134  & 1.86  \\
     \texttt{mk-en} & 190,493  & 19,747  & 17,747  & 2,000 & 13,134  & 3.06  \\
    \hline
     %\texttt{eu-es} & 765,115  & 10,190,079  & 87,907  & 85,907   & 1,000 & 1,000 & 1,806  & 1.30 \\
     \texttt{eu-es} & 765,115  & 87,907  & 85,907  & 2,000 & 1,806  & 3.11 \\
    \hline
     %\texttt{en-es} & 1,467,708 & 30,154,098  & 312,162 & 310,162 & 1,000 & 1,000 & 2,082  & 1.08 \\
     \texttt{en-es} & 1,467,708 & 312,162 & 310,162 & 2,000 & 2,082  & 2.28 \\
    \hline  
 \end{tabular}
 \caption{Statistics about the source corpora. The column {\bf other}
   gives the number of sentences reserved for testing (1,000) and
   development (1,000). The column {\bf no.\ amb} gives the number of
   unique tokens with more than one possible translation. The column
   {\bf av.\ amb} gives the average number of translations per ambiguous
   word. This is calculated by looking up each word in the corpus in
   the bilingual dictionary of the MT system and dividing the total
   number of translation by the number of words. Both {\bf av.\ amb}
   and {\bf no.\ amb} are calculated over the whole corpus.}
 \label{table:input-corp}
\end{center}
\end{table}

@@

\begin{table}
\begin{center}
 \begin{tabular}{|l|r|r|r|r|r|}
    \hline 
    {\bf Pair}      & {\bf Lines} & {\bf SL words} & {\bf TL words} & {\bf Amb. words} & {\bf \% ambig} \\ 
    \hline 
     \texttt{br-fr} & 1,000 & 13,854  & 13,878  & 1,163 & 8.39   \\
    \hline 
     \texttt{mk-en} & 1,000 & 13,441 & 14,228  & 3,872 & 28.80 \\
    \hline 
     \texttt{eu-es} & 1,000 & 7,967 & 11,476  & 1,360  & 17.07 \\
    \hline 
     \texttt{en-es} & 1,000 & 19,882 & 20,944  & 1,469  & 7.38 \\
    \hline 
 \end{tabular} 
\end{center}
 \caption{Statistics about the test corpora. The columns \textbf{amb.\ words} and \textbf{\%~ambig} gives 
    the number of word with more than one translation and the percentage of SL words which have 
    more than one translation respectively.}
  \label{table:test-corp}
\end{table}

@@

\begin{table}
 \begin{center}

@@

 \end{center}
 \caption{LER and BLEU scores with 95\% confidence intervals for the reference
   systems on the test corpora. The \texttt{max-ent} system has been trained using 
   fractional counts. The results in bold face show statistically significant
   improvements for the maximum-entropy model compared to the target language model 
   according to pair-bootstrap resampling.}
  \label{table:eval-mono-results}
\end{table}

>>>>

\author{
 Francis M.\ Tyers \\
 HSL-fakultehta, \\
 UiT Norgga \'{a}rktala\v{s} universitehta, \\
 N-9018 Romsa \\
 \And
 Felipe S\'anchez-Mart\'inez \\
 Dept.\ Lleng.\ i Sist.\ Inform., \\
 Universitat d'Alacant,\\
 E-03071 Alacant\\
 \And
 Mikel L.\ Forcada \\
 Dept.\ Lleng.\ i Sist.\ Inform., \\
 Universitat d'Alacant,\\
 E-03071 Alacant\\
}

@@

\begin{table*}
 \begin{center}
  \begin{tabular}{|c|l|r|}
    \hline
       & \textbf{Sentence} & $p(g_i|S)$ \\
    \hline % Los pescados grandes peque\~{n}o come
     \(S_1\) & \emph{Arrain handiak txikia jaten du}  &  \\
                        $\tau(g_{11},S_1)$ & El pez grande se come el peque\~{n}o &  0.830 \\
                        $\tau(g_{12},S_1)$ & El pescado grande se come el peque\~{n}o & 0.134 \\
                        $\tau(g_{13},S_1)$ & El pescado capaz se come el peque\~{n}o & 0.034\\
                        $\tau(g_{14},S_1)$ & El pez capaz se come el peque\~{n}o  & 0.001\\
    \hline % El padre el pescado nos ha preparado para cenar
     \(S_2\) & \emph{Aitak arraina prestatu digu afaltzeko} &  \\
                        $\tau(g_{21},S_2)$ & Padre nos ha preparado pescado para cenar & 0.846 \\
                        $\tau(g_{22},S_2)$ & Pap\'{a} nos ha preparado pescado para cenar & 0.134 \\
                        $\tau(g_{23},S_2)$ & Padre nos ha preparado pez para cenar & 0.015 \\
                        $\tau(g_{24},S_2)$ & Pap\'{a} nos ha preparado pez para cenar & 0.005 \\
    \hline % Aqu\'{\i} el pescado muy es dulce
    $S_3$  & \emph{Hemen arraina oso goxoa da}  & \\
                        $\tau(g_{31},S_3)$ & Aqu\'{\i} el pescado es muy rico & 0.912 \\
                        $\tau(g_{32},S_3)$ & Aqu\'{\i} el pescado es muy suave & 0.067 \\
                        $\tau(g_{33},S_3)$ & Aqu\'{\i} el pez es muy rico & 0.011 \\
                        $\tau(g_{34},S_3)$ & Aqu\'{\i} el pez es muy suave & 0.008 \\
                        $\tau(g_{35},S_3)$ & Aqu\'{\i} el pez es muy dulce & 0.001 \\
                        $\tau(g_{36},S_3)$ & Aqu\'{\i} el pescado es muy dulce & 0.001 \\
    \hline % Un pescado grande del mar es
    \(S_4\) & \emph{Itsasoko arrain handi bat da} & \\
                        $\tau(g_{41},S_4)$ & Es un pez grande del mar & 0.595 \\
                        $\tau(g_{42},S_4)$ & Es un pescado grande del mar & 0.403 \\
                        $\tau(g_{43},S_4)$ & Es un pez capaz del mar & 0.001 \\
                        $\tau(g_{44},S_4)$ & Es un pescado capaz del mar & 0.001 \\
    \hline % El pescado grande por esos serv\'{\i}an para alimentar
    $S_5$ &  \emph{Arrain handiak horiez baliatzen ziren elikatzeko} &  \\
                        $\tau(g_{51},S_5)$ & Los peces grandes se nutr\'{\i}an de esos & 0.862 \\
                        $\tau(g_{52},S_5)$ & Los pescados grandes se nutr\'{\i}an de esos & 0.121 \\
                        $\tau(g_{53},S_5)$ & Los peces grandes se alimentaban de esos & 0.012 \\
                        $\tau(g_{54},S_5)$ & Los peces capaces se alimentaban de esos & 0.001 \\
                        $\tau(g_{55},S_5)$ & Los pescados grandes se alimentaban de esos & 0.001 \\
                        $\tau(g_{56},S_5)$ & Los pescados capaces se alimentaban de esos & 0.001 \\
                        $\tau(g_{57},S_6)$ & Los peces capaces se nutr\'{\i}an de esos & 0.001 \\
                        $\tau(g_{58},S_7)$ & Los pescados capaces se nutr\'{\i}an de esos & 0.001 \\
    \hline

  \end{tabular}
  \caption{A sample of source-language sentences, the output of the
    possible lexical-selection paths after scoring on a
    target-language model. The scores are normalised as fractional
    counts, which are used as inputs to the maximum-entropy training
    process. In all sentences, the resultant translation with the highest share of the 
    probability mass is the most adequate.}
   \label{tb:workedexample}
 \end{center}
\end{table*}

@@

\begin{table*}
  \centering
  \begin{tabular}{|l|r|r|r|r|r|}
    \hline
    \multirow{2}{*}{\textbf{Pair}} & \textbf{Lexical} & \multirow{2}{*}{\textbf{Ambig.}} & \textbf{Na\"{i}ve}& \textbf{Total} & \textbf{Ambig. of}\\\
                          & \textbf{items} &                            & \textbf{coverage} & \textbf{ambig.} & \textbf{ambig. words}\\
    \hline
    \texttt{br-fr} & 27,690 & 235  &  94.47 \%  & 1.02 & 2.5 \\
    \texttt{mk-en} & 33,229 & 1,323 & 92.17 \%  & 1.37 & 2.8 \\
    \texttt{eu-es} & 19,323 & 455 & 91.70 \% & 1.08 & 2.4 \\
    \texttt{en-es} & 30,186 & 822 & 98.08 \% & 1.05 & 2.2 \\
    \hline
  \end{tabular}
  \caption{Statistics about development status of language pairs. The second
    column gives the number of entries in the bilingual dictionaries of the systems;
    the third column gives the total number of source language words with more 
    than one translation in the target language; the fourth column gives the coverage
    of the dictionaries on the test corpus (see section~\ref{sec:eval-corpus}); the fifth
    column gives the total mean ambiguity for open classes (nouns, verbs, adjectives);
    and the final column gives the mean ambiguity for ambiguous words in open classes.}
  \label{table:lang-pair-stats}
\end{table*}

@@

\label{sec:eval-metrics}
\subsubsection{Lexical-selection performance}
This is an intrinsic module-based evaluation of the performance of the
lexical-selection module.  It measures how well the lexical-selection
module disambiguates the output of the lexical-transfer module as
compared to a gold-standard corpus. For this task, we define a metric,
the lexical-selection error rate ({\sc ler}),
% see section~\ref{sec:ler}. 
\begin{figure*}
\begin{center}
\setlength\tabcolsep{5.2pt}%
 \begin{tabular}{|l|llllll|}
   \hline
   \multicolumn{7}{|c|}{L'estiu \'{e}s una estaci\'{o} llarga}\\
   \hline
   $S$        & \emph{el} & \emph{estiu} & \emph{ser} & \emph{un} & \emph{estaci\'{o}} & \emph{llarg} \\
   \hline
   $T_s(s_i)$ & \{the\} & \{summer\} & \{be\} & \{a\} & \{station, season\} &  \{long, lengthy\} \\
   \hline 
   $T_r(s_i)$ & \{the\} & \{summer\} & \{be\} & \{a\} &  \{season\} & \{long\} \\
   \hline
   $T_t(s_i)$ & \{the\} & \{summer\} & \{be\} & \{a\} &  \{station\} & \{long\} \\
   \hline
 \end{tabular}
 \caption{An example input sentence in Catalan and the three sets of
   English translations used for calculating the lexical-selection
   error rate. The source sentence $S = (s_1, s_2, \ldots, s_{|S|})$
   has two ambiguous words, \emph{estaci\'{o}} and \emph{llarg}. There is one difference
   between the reference set $T_r(s_i)$ and the test set $T_t(s_i)$ of
   translations; as a result, the error rate for this sentence is
   50\%.}
\label{fig:ler-example}
\end{center}
\end{figure*}

@@

\begin{table*}
\begin{center}
 \begin{tabular}{|l|r|r||r|r|r|r|}
    \hline  
    %Pair           & Lines & Words & Extracted & \texttt{train} & \texttt{test} & \texttt{dev} & No. ambig & Av. ambig \\ 
    {\bf Pair}      & {\bf Lines} & {\bf Extract.} & \texttt{train} & \texttt{other} & {\bf No. amb} & {\bf Av. amb} \\ 
    \hline  
     %\texttt{br-fr} & 57,305  & 702,328  & 4,668 & 2,668 & 1,000  & 1,000 & 603  & 1.07 \\
     \texttt{br-fr} & 57,305  & 4,668 & 2,668 & 2,000  & 603  & 3.06 \\
    \hline  
     %\texttt{mk-en} & 190,493  & 4,259,338  & 19,747  & 17,747  & 1,000 & 1,000 & 13,134  & 1.86  \\
     \texttt{mk-en} & 190,493  & 19,747  & 17,747  & 2,000 & 13,134  & 3.06  \\
    \hline
     %\texttt{eu-es} & 765,115  & 10,190,079  & 87,907  & 85,907   & 1,000 & 1,000 & 1,806  & 1.30 \\
     \texttt{eu-es} & 765,115  & 87,907  & 85,907  & 2,000 & 1,806  & 3.11 \\
    \hline
     %\texttt{en-es} & 1,467,708 & 30,154,098  & 312,162 & 310,162 & 1,000 & 1,000 & 2,082  & 1.08 \\
     \texttt{en-es} & 1,467,708 & 312,162 & 310,162 & 2,000 & 2,082  & 2.28 \\
    \hline  
 \end{tabular}
 \caption{Statistics about the source corpora. The column {\bf other}
   gives the number of sentences reserved for testing (1,000) and
   development (1,000). The column {\bf no.\ amb} gives the number of
   unique tokens with more than one possible translation. The column
   {\bf av.\ amb} gives the average number of translations per ambiguous
   word. This is calculated by looking up each word in the corpus in
   the bilingual dictionary of the MT system and dividing the total
   number of translation by the number of words. Both {\bf av.\ amb}
   and {\bf no.\ amb} are calculated over the whole corpus.}
 \label{table:input-corp}
\end{center}
\end{table*}

@@

\begin{table*}
\begin{center}
 \begin{tabular}{|l|r|r|r|r|r|}
    \hline 
    {\bf Pair}      & {\bf Lines} & {\bf SL words} & {\bf TL words} & {\bf Amb. words} & {\bf \% ambig} \\ 
    \hline 
     \texttt{br-fr} & 1,000 & 13,854  & 13,878  & 1,163 & 8.39   \\
    \hline 
     \texttt{mk-en} & 1,000 & 13,441 & 14,228  & 3,872 & 28.80 \\
    \hline 
     \texttt{eu-es} & 1,000 & 7,967 & 11,476  & 1,360  & 17.07 \\
    \hline 
     \texttt{en-es} & 1,000 & 19,882 & 20,944  & 1,469  & 7.38 \\
    \hline 
 \end{tabular} 
\end{center}
 \caption{Statistics about the test corpora. The columns \textbf{amb.\ words} and \textbf{\%~ambig} gives 
    the number of word with more than one translation and the percentage of SL words which have 
    more than one translation respectively.}
  \label{table:test-corp}
\end{table*}

@@

\begin{table*}
 \begin{center}

@@

 \end{center}
 \caption{LER and BLEU scores with 95\% confidence intervals for the reference
   systems on the test corpora. The \texttt{max-ent} system has been trained using 
   fractional counts. The results in bold face show statistically significant
   improvements for the maximum-entropy model compared to the target language model 
   according to pair-bootstrap resampling.}
  \label{table:eval-mono-results}
\end{table*}
